Manuscript draft

### 2019-06-10 manuscript/manuscript draft.Rmd
updated manuscript draft

---
title: "Manuscript draft"
output:
  pdf_document:
    fig_caption: yes
bibliography: library.bib
csl: global-change-biology.csl
header-includes:  \usepackage{float}
---

@@

Globally, forests play an important role in the carbon cycle and are an important component of global carbon dioxide budgets [@Luyssaert2008]. They show higher levels of productivity than non-forest  terrestrial ecosystems [@DelGrosso2008], and as a result achieve significant carbon sequestration and storage. Estimating the total role of forests in the carbon cycle is challenging, but studies indicate that old growth forests alone sequester up to 1.4 GtCyr^-1^ [@Malhi1999], while the total sequestration of carbon by established forests globally could be up to 2.4 GtCyr^-1^, with the largest sinks being in old-growth tropical forests [@Pan2011]. As atmospheric carbon dioxide levels continue to rise, with consequences for global climate, there is increasing recognition that proper protection and management of forest resources will have an important role to play in mitigating climate change. Understanding the patterns of forest productivity on a global scale, and the drivers behind them, is therefore a priority in forest research.

There are two major questions to understand: firstly, how forest productivity varies globally - and specifically which areas show the greatest peaks in productivity -; and secondly, which climate variables drive this variation. On a global scale, the productivity of forests varies with latitude, with a general trend of increasing productivity towards the tropics [@Beer2010; @Jung2011]; however the exact nature of this pattern, and how it varies by component of productivity, is poorly understood. This latitudinal gradient is most likely to be explained by climatic gradients in temperature, precipitation, length of growing season, and combinations of the above. Productivity is influenced by a range of climatic drivers, including mean annual temperature (MAT) and mean annual precipitation (MAP) [@DelGrosso2008], but doesn't necessarily respond linearly to these drivers. Disentangling the shape of productivity responses to climate drivers will enable better predictions of future responses under climate change.

@@

Currently, GPP cannot be measured directly by observing total ecosystem photosynthesis. Instead field estimates of GPP have to be derived based on modelling and extrapolation of eddy-covariance studies and measurements of net ecosystem exchange (NEE) [@Clark2017].

@@

$$ ANPP = NPP~stem~ + NPP~branch~ + NPP~canopy~ $$
where NPP~stem~ is the annual woody increment of all stems above a specified diameter at breast height (DBH), NPP~branch~ is annual woody branch turnover, and NPP~canopy~ is annual foliage production, including leaves, twigs, and reproductive structures. ANPP may also include NPP~VOC~, the annual emission of volatile organic compounds, and NPP~herbivory~, the annual consumption of plant matter by herbivores, but these components are often excluded from field observations as they are much harder to quantify. Other components of aboveground productivity that remain largely unquantified include epiphytes, hemiepiphytes, and understory plants [@Clark2017]. All current ANPP estimates are based on the assumption that the contribution of these components to overall NPP is insignificant.

@@

$$ BNPP = NPP~coarse root~ + NPP~fine root~ $$
where NPP~coarse root~ is the annual production of coarse roots (typically roots >2mm diameter), and NPP~fine root~ is the annual production of fine roots (typically roots <2mm diameter) [@Aragao2009]. Calculations of BNPP may also include NPP~exudation~, a measure of annual carbon losses through root exudation, and NPP~symbionts~, the annual carbon allocation to mycorrhizae and legumes, but, as before, this is challenging to quantify and is often excluded from field observations.

BNPP is a poorly understood component of total ecosystem productivity, primarily because of the challenges in gaining accurate field measurements. Coarse root productivity is often estimated via extrapolation of NPP~stem~ estimates using allometries that may not have been empirically verified [@Clark2001]. NPP~fine root~ is easier to quantify through soil cores and minirhizotrons, however, sampling tends to be limited to the surface soils, with very few studies sampling to depths below 3 metres [cite]. As a result, it is possible that BNPP is currently significantly underestimated, despite being a hugely significant component of total ecosystem productivity [@Pan2011].

*Which factors influence primary productivity?* Primary productivity is influenced by many factors, which often act across a range of scales, and may show interactive effects with each other. On a local scale, stand age [@Litton2007; @Gillman2015], management [@Simova2017]; nutrient availability [@Aragao2009]; and altitude [@Girardin2010; @Malhi2017] all impact forest productivity. On a global scale, changes in primary productivity are influenced by climatic variables and abiotic gradients, such as the length of growing season [@Michaletz2014]. There is some debate over the precise relationship between these drivers and productivity; While mean annual precipitation (MAP) and mean annual temperature (MAT) have been argued to be significant predictors of productivity [@Chu2016], other studies have found that the correlation between productivity and MAT is a factor of the relationship between productivity and growing season length [@Kerkhoff2005; @Malhi2012; @Michaletz2014; @Michaletz2018]. Improving understanding of how these factors interact to control global patterns in primary productivity is essential to understanding the global carbon cycle.

Current research into how primary productivity varies with latitude is inconclusive, and - though it has primarily focussed on patterns of GPP, NPP, and ANPP - indicates that different components of productivity may show different relationships to latitude. Gross primary productivity is generally thought to be highest in the tropics. Modelling of global terrestrial ecosystem GPP though upscaling and calibration of eddy flux measurements indicates a peak in GPP in the tropics, with the highest levels in tropical forests [@Beer2010; @Jung2011]. This is corroborated by analysis of site-level GPP measurements, which show a strong positive correlation between GPP and MAT and MAP [@Luyssaert2007], with the highest GPP values reported in tropical forests. The influence of latitude on global patterns of NPP is less clear than that of GPP. Simova and Storch (2017) found that, as with GPP, NPP decreases with latitude, peaking in the tropics. However, other studies have found the highest values of NPP in temperate forests [@Luyssaert2007; @Huston2009]. Because of the challenges in accurately obtaining unbiased measures of belowground productivity, many studies focus on ANPP in preference to measures of NPP. Studies on global patterns of ANPP are equally inconclusive: Gillman et al. [-@Gillman2015] found a weak negative relationship between ANPP and latitude, with the relationship becoming stronger in older forest stands. These findings were echoed in other studies, which have found weak or no relationships between ANPP and latitude [@Huston2009].

Furthermore, there is evidence that different components of productivity show individual responses to drivers of productivity. For example, increases in GPP have been reported to saturate above 25$^\circ$C MAT [@Larjavaara2012], while increases in NPP are recorded to saturate above 10$^\circ$C MAT [@Luyssaert2007]. Similarly, allocation to different components of ANPP varies with climate. Within the tropics, allocation to canopy NPP appears fairly consistent, with significantly greater variation in allocation to woody and belowground NPP [@Litton2007; @Malhi2011]. Allocation to these structural biomass components has been shown to increase with water availability [@Litton2007; @Bloom2016], and is highest in the wet tropics, indicating that control of woody productivity by MAP may be more significant than control of foliar productivity. However, these studies are regional, meaning that our understanding of variation in allocation and its relationship to climate on a global scale remains limited. In addition, allocation is also influenced by stand age [@DeLucia2007], nutrient availability [@Litton2007] and forest structure [@Taylor2019], which can make it challenging to disentangle the effects of climate.

@@

Analyses were conducted on data contained in the open-access ForC database [@Anderson-Teixeira2016; @Anderson-Teixeira2018]. This database contains records of field-based measurements of forest carbon stocks and annual fluxes, compiled from original publications and existing data compilations and databases. Associated data, such as stand age, measurement methodologies, and disturbance history, are also included. For each site, site geographic co-ordinates were used to extract the Koeppen-Geiger zone from the ESRI Koeppen-Geiger map [cite], and the FAO Global Ecological Zone from the ???. Additional targeted literature searches were conducted to identify any further available data on primary productivity, with particular focus on old-growth forests in temperate and boreal regions. ForC currently contains [] records from [] plots, representing [] distinct geographic areas across all forested biogeographic and climate zones.

@@

A subset of the ForC database was generated for the purposes of this analysis, in order to control for data quality and remove biasing factors. Since management can alter observed patterns of primary productivity [@Simova2017], sites were excluded from analysis if they were managed, defined as plots that were planted, managed as plantations, irrigated, fertilised or including the term "managed" in their site description. Sites that had experienced significant disturbance were also excluded. Disturbances that justified site exclusion were major cutting or harvesting, and/or burning, flooding, drought and storm events with site mortality >10% of trees. Grazed sites were retained.  
There is evidence that stand age influences patterns of primary productivity and carbon allocation in forest ecosystems, and can confound relationships between latitude and primary productivity [@DeLucia2007; @Gillman2015]. To reduce any biasing effects of stand age, stands under 100 years of age were excluded from analysis. Sites for which stand age was unknown were excluded from analysis.

@@

| Database | Variables downloaded | Citation |
|-----------------------|--------------------------------------------|---------------------|
| WorldClim | Mean annual temperature; temperature seasonality; annual temperature range; mean annual precipitation | [@Hijmans2005] |
| WorldClim2 | Vapour pressure; solar radiation | [@Fick2017] |
| Climate Research Unit (CRU) time-series dataset v 4.03 | Cloud cover; annual frost days; annual wet days; potential evapotranspiration | [@Harris2014] |
| Global Aridity Index and Potential Evapotranspiration Climate Database | Aridity; potential evapotranspiration | [@Trabucco2018] |
| TerraClimate | Vapour pressure deficit | [@Abatzoglou2018] |

@@

*Model specification.* The effects of climate and latitude on primary productivity were analysed using mixed effects models using the package 'lme4' in RStudio v. 1.1.463 (cite). The effect of each extracted climate variable on each measure of primary productivity was modelled by specifying the climate variable as a fixed effect. Site altitude was also included as a fixed effect. Random effect was stand nested within area. Data from the temperate regions was heavily skewed towards studies from the old-growth forests of the Pacific Northwest. These forests have very high productivity, and so to reduce any bias from over-sampling of this region, the proportion of global forest cover contributed by each Koeppen climate zone was calculated, and the models weighted according these proportions.

>>>>

---
title: "Manuscript draft"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: library.bib
biblio-style: apalike
csl: apa.csl
header-includes:  \usepackage{float}

---

@@

Globally, forests play an important role in the carbon cycle and are an important component of global carbon dioxide budgets [@luyssaert_old-growth_2008]. They show higher levels of productivity than non-forest  terrestrial ecosystems [@del_grosso_global_2008], and as a result achieve significant carbon sequestration and storage. Estimating the total role of forests in the carbon cycle is challenging, but studies indicate that old growth forests alone sequester up to 1.4 GtCyr^-1^ [@malhi_carbon_1999], while the total sequestration of carbon by established forests globally could be up to 2.4 GtCyr^-1^, with the largest sinks being in old-growth tropical forests [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, with consequences for global climate, there is increasing recognition that proper protection and management of forest resources will have an important role to play in mitigating climate change. Understanding the patterns of forest productivity on a global scale, and the drivers behind them, is therefore a priority in forest research.

There are two major questions to understand: firstly, how forest productivity varies globally - and specifically which areas show the greatest peaks in productivity -; and secondly, which climate variables drive this variation. On a global scale, the productivity of forests varies with latitude, with a general trend of increasing productivity towards the tropics [@beer_terrestrial_2010; @jung_global_2011]; however the exact nature of this pattern, and how it varies by component of productivity, is poorly understood. This latitudinal gradient is most likely to be explained by climatic gradients in temperature, precipitation, length of growing season, and combinations of the above. Productivity is influenced by a range of climatic drivers, including mean annual temperature (MAT) and mean annual precipitation (MAP) [@del_grosso_global_2008], but doesn't necessarily respond linearly to these drivers. Disentangling the shape of productivity responses to climate drivers will enable better predictions of future responses under climate change.

@@

Currently, GPP cannot be measured directly by observing total ecosystem photosynthesis. Instead field estimates of GPP have to be derived based on modelling and extrapolation of eddy-covariance studies and measurements of net ecosystem exchange (NEE) [@clark_reviews_2017].

@@

$$ ANPP = NPP~stem~ + NPP~branch~ + NPP~canopy~ $$
where NPP~stem~ is the annual woody increment of all stems above a specified diameter at breast height (DBH), NPP~branch~ is annual woody branch turnover, and NPP~canopy~ is annual foliage production, including leaves, twigs, and reproductive structures. ANPP may also include NPP~VOC~, the annual emission of volatile organic compounds, and NPP~herbivory~, the annual consumption of plant matter by herbivores, but these components are often excluded from field observations as they are much harder to quantify. Other components of aboveground productivity that remain largely unquantified include epiphytes, hemiepiphytes, and understory plants [@clark_reviews_2017]. All current ANPP estimates are based on the assumption that the contribution of these components to overall NPP is insignificant.

@@

$$ BNPP = NPP~coarse root~ + NPP~fine root~ $$
where NPP~coarse root~ is the annual production of coarse roots (typically roots >2mm diameter), and NPP~fine root~ is the annual production of fine roots (typically roots <2mm diameter) [@aragao_above-_2009]. Calculations of BNPP may also include NPP~exudation~, a measure of annual carbon losses through root exudation, and NPP~symbionts~, the annual carbon allocation to mycorrhizae and legumes, but, as before, this is challenging to quantify and is often excluded from field observations.

BNPP is a poorly understood component of total ecosystem productivity, primarily because of the challenges in gaining accurate field measurements. Coarse root productivity is often estimated via extrapolation of NPP~stem~ estimates using allometries that may not have been empirically verified [@clark_net_2001]. NPP~fine root~ is easier to quantify through soil cores and minirhizotrons, however, sampling tends to be limited to the surface soils, with very few studies sampling to depths below 3 metres [cite]. As a result, it is possible that BNPP is currently significantly underestimated, despite being a hugely significant component of total ecosystem productivity [@pan_large_2011].

*Which factors influence primary productivity?* Primary productivity is influenced by many factors, which often act across a range of scales, and may show interactive effects with each other. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], management [@simova_enigma_2017]; nutrient availability [@aragao_above-_2009]; and altitude [@girardin_net_2010; @malhi_variation_2017] all impact forest productivity. On a global scale, changes in primary productivity are influenced by climatic variables and abiotic gradients, such as the length of growing season [@michaletz_convergence_2014]. There is some debate over the precise relationship between these drivers and productivity; While mean annual precipitation (MAP) and mean annual temperature (MAT) have been argued to be significant predictors of productivity [@chu_does_2016], other studies have found that the correlation between productivity and MAT is a factor of the relationship between productivity and growing season length [@kerkhoff_plant_2005; @malhi_productivity_2012; @michaletz_convergence_2014; @michaletz_drivers_2018]. Improving understanding of how these factors interact to control global patterns in primary productivity is essential to understanding the global carbon cycle.

Current research into how primary productivity varies with latitude is inconclusive, and - though it has primarily focussed on patterns of GPP, NPP, and ANPP - indicates that different components of productivity may show different relationships to latitude. Gross primary productivity is generally thought to be highest in the tropics. Modelling of global terrestrial ecosystem GPP though upscaling and calibration of eddy flux measurements indicates a peak in GPP in the tropics, with the highest levels in tropical forests [@beer_terrestrial_2010; @jung_global_2011]. This is corroborated by analysis of site-level GPP measurements, which show a strong positive correlation between GPP and MAT and MAP [@luyssaert_co2_2007], with the highest GPP values reported in tropical forests. The influence of latitude on global patterns of NPP is less clear than that of GPP. Simova and Storch (2017) found that, as with GPP, NPP decreases with latitude, peaking in the tropics. However, other studies have found the highest values of NPP in temperate forests [@luyssaert_co2_2007; @huston_global_2009]. Because of the challenges in accurately obtaining unbiased measures of belowground productivity, many studies focus on ANPP in preference to measures of NPP. Studies on global patterns of ANPP are equally inconclusive: Gillman et al. (2015) found a weak negative relationship between ANPP and latitude, with the relationship becoming stronger in older forest stands. These findings were echoed in other studies, which have found weak or no relationships between ANPP and latitude [@huston_global_2009].

Furthermore, there is evidence that different components of productivity show individual responses to drivers of productivity. For example, increases in GPP have been reported to saturate above 25$^\circ$C MAT [@larjavaara_temperature_2012], while increases in NPP are recorded to saturate above 10$^\circ$C MAT [@luyssaert_co2_2007]. Similarly, allocation to different components of ANPP varies with climate. Within the tropics, allocation to canopy NPP appears fairly consistent, with significantly greater variation in allocation to woody and belowground NPP [@litton_carbon_2007; @malhi_allocation_2011]. Allocation to these structural biomass components has been shown to increase with water availability [@litton_carbon_2007; @bloom_decadal_2016], and is highest in the wet tropics, indicating that control of woody productivity by MAP may be more significant than control of foliar productivity. However, these studies are regional, meaning that our understanding of variation in allocation and its relationship to climate on a global scale remains limited. In addition, allocation is also influenced by stand age [@de_lucia_forest_2007], nutrient availability [@litton_carbon_2007] and forest structure [@taylor_greater_2019], which can make it challenging to disentangle the effects of climate.

@@

Analyses were conducted on data contained in the open-access ForC database [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018]. This database contains records of field-based measurements of forest carbon stocks and annual fluxes, compiled from original publications and existing data compilations and databases. Associated data, such as stand age, measurement methodologies, and disturbance history, are also included. For each site, site geographic co-ordinates were used to extract the Koeppen-Geiger zone from the ESRI Koeppen-Geiger map [cite], and the FAO Global Ecological Zone from the ???. Additional targeted literature searches were conducted to identify any further available data on primary productivity, with particular focus on old-growth forests in temperate and boreal regions. ForC currently contains [] records from [] plots, representing [] distinct geographic areas across all forested biogeographic and climate zones.

@@

A subset of the ForC database was generated for the purposes of this analysis, in order to control for data quality and remove biasing factors. Since management can alter observed patterns of primary productivity [@simova_enigma_2017], sites were excluded from analysis if they were managed, defined as plots that were planted, managed as plantations, irrigated, fertilised or including the term "managed" in their site description. Sites that had experienced significant disturbance were also excluded. Disturbances that justified site exclusion were major cutting or harvesting, and/or burning, flooding, drought and storm events with site mortality >10% of trees. Grazed sites were retained.  
There is evidence that stand age influences patterns of primary productivity and carbon allocation in forest ecosystems, and can confound relationships between latitude and primary productivity [@de_lucia_forest_2007; @gillman_latitude_2015]. To reduce any biasing effects of stand age, stands under 100 years of age were excluded from analysis. Sites for which stand age was unknown were excluded from analysis.

@@

| Database | Variables downloaded | Citation |
|-----------------------|--------------------------------------------|---------------------|
| WorldClim | Mean annual temperature; temperature seasonality; annual temperature range; mean annual precipitation | [@hijmans_very_2005] |
| WorldClim2 | Vapour pressure; solar radiation | [@fick_worldclim_2017] |
| Climate Research Unit (CRU) time-series dataset v 4.03 | Cloud cover; annual frost days; annual wet days; potential evapotranspiration | [@harris_updated_2014] |
| Global Aridity Index and Potential Evapotranspiration Climate Database | Aridity; potential evapotranspiration | [@trabucco_global_2018] |
| TerraClimate | Vapour pressure deficit | [@abatzoglou_terraclimate_2018] |

@@

*Model specification.* The effects of climate and latitude on primary productivity were analysed using mixed effects models using the package 'lme4' [@bates_fitting_2015] in R v.3.5.1 [@r_core_team_r_2018]. The effect of each extracted climate variable on each measure of primary productivity was modelled by specifying the climate variable as a fixed effect. Site altitude was also included as a fixed effect. Random effect was stand nested within area. Data from the temperate regions was heavily skewed towards studies from the old-growth forests of the Pacific Northwest. These forests have very high productivity, and so to reduce any bias from over-sampling of this region, the proportion of global forest cover contributed by each Koeppen climate zone was calculated, and the models weighted according these proportions.
